---
title: "LSTM Forecast: Daily Average Uber Trip Duration"
author: "Farooq Mahmud"
format: pdf
editor: source
---

## Attribution

- **Data and structure:** Daily average trip duration series and overall workflow follow the STA6856 final project ([finalproject.qmd](https://github.com/farooq-teqniqly/uwf-sta6856), `finalproject/code/finalproject.qmd`). Data source: NYC TLC High-Volume FHV data; aggregated to daily averages as in that project.
- **Plot style:** ggplot2 layout, `theme_minimal()`, and `scale_x_date()` usage are aligned with [finalproject.qmd](https://github.com/farooq-teqniqly/uwf-sta6856) for consistency with the ARIMA report.
- **LSTM implementation:** LSTM design (sequence construction, scaling, recursive forecast) follows common Keras time-series patterns (see [Keras timeseries documentation](https://keras.io/examples/timeseries/)); R interface via the [keras](https://keras.rstudio.com/) package.
- **Evaluation metrics:** sMAPE and MASE definitions follow forecasting comparison literature (e.g. Makridakis et al.; Prabhat et al. as reviewed in the capstone hw3 literature review).

## Load the data

```{r}
data <- read.csv(file.path(getwd(), "../data", "finalproject.csv"))
data$pickup_date <- as.Date(data$pickup_date)
data <- data[order(data$pickup_date), ]
rownames(data) <- NULL
```

## Preview data

```{r}
dim(data)
head(data, 10)
summary(data$avg_duration_min)
sum(is.na(data))
```

## Preprocessing

### Train / validation / test split (chronological)

```{r}
HORIZON <- 14   # 14-day forecast (same as ARIMA)
VAL_DAYS <- 30
n <- nrow(data)
test_end <- n
test_start <- test_end - HORIZON
val_end <- test_start
val_start <- val_end - VAL_DAYS
train_end <- val_start

values <- matrix(data$avg_duration_min, ncol = 1)
train_values <- values[1:train_end, , drop = FALSE]
val_values <- values[val_start:(val_end - 1), , drop = FALSE]
test_values <- values[test_start:(test_end - 1), , drop = FALSE]
test_dates <- data$pickup_date[test_start:(test_end - 1)]

cat("Train: 1:", train_end, ", Val: ", val_start, ":", val_end - 1, ", Test: ", test_start, ":", test_end - 1, "\n")
cat("Train size:", nrow(train_values), ", Val:", nrow(val_values), ", Test:", nrow(test_values), "\n")
```

### Scaling

```{r}
min_val <- min(train_values)
max_val <- max(train_values)
scale_fun <- function(x) (x - min_val) / (max_val - min_val)
inv_scale_fun <- function(z) z * (max_val - min_val) + min_val

train_scaled <- scale_fun(train_values)
val_scaled <- scale_fun(val_values)
test_scaled <- scale_fun(test_values)
```

### Sequence construction (sliding window)

```{r}
SEQ_LEN <- 21

create_sequences <- function(scaled_data, seq_len) {
  X <- array(NA, dim = c(nrow(scaled_data) - seq_len, seq_len, 1))
  y <- numeric(nrow(scaled_data) - seq_len)
  for (i in seq_len(nrow(scaled_data) - seq_len)) {
    ii <- i + seq_len - 1
    X[i, , 1] <- scaled_data[i:ii, 1]
    y[i] <- scaled_data[ii + 1, 1]
  }
  list(X = X, y = y)
}

train_val_scaled <- rbind(train_scaled, val_scaled)
seqs <- create_sequences(train_val_scaled, SEQ_LEN)
X_train <- seqs$X
y_train <- seqs$y
cat("X_train dim:", dim(X_train), ", y_train length:", length(y_train), "\n")
```

## Model (LSTM)

```{r}
library(keras)

# Functional API (compatible with Keras 3 / TensorFlow 2.16+)
inputs <- layer_input(shape = c(SEQ_LEN, 1))
outputs <- inputs %>%
  layer_lstm(units = 32, return_sequences = FALSE) %>%
  layer_dropout(rate = 0.2) %>%
  layer_dense(units = 1)
model <- keras_model(inputs, outputs)

model$compile(
  optimizer = optimizer_adam(),
  loss = "mse"
)
summary(model)
```

## Training

```{r}
library(tensorflow)
# Use 20% of training sequences as validation for early stopping
val_idx <- sample(length(y_train), size = round(0.2 * length(y_train)))
x_val <- X_train[val_idx, , , drop = FALSE]
y_val <- y_train[val_idx]
x_fit <- X_train[-val_idx, , , drop = FALSE]
y_fit <- y_train[-val_idx]

# Keras 3 expects tf.Tensor with int32-compatible shapes; use numpy to avoid R→float shape
np <- reticulate::import("numpy")
x_fit_tf <- tf$constant(np$array(x_fit, dtype = np$float32))
y_fit_tf <- tf$constant(np$array(y_fit, dtype = np$float32))
x_val_tf <- tf$constant(np$array(x_val, dtype = np$float32))
y_val_tf <- tf$constant(np$array(y_val, dtype = np$float32))

history <- model$fit(
  x_fit_tf, y_fit_tf,
  epochs = 100L,
  batch_size = 16L,
  validation_data = list(x_val_tf, y_val_tf),
  callbacks = list(
    callback_early_stopping(monitor = "val_loss", patience = 15, restore_best_weights = TRUE)
  ),
  verbose = 1
)
```

## Training and validation loss (ggplot2)

```{r}
library(ggplot2)
library(tidyr)

# Keras 3 History uses .history (dict of metric name -> list of values per epoch)
hist_df <- as.data.frame(history$history)
hist_df$epoch <- seq_len(nrow(hist_df))
hist_long <- pivot_longer(hist_df, cols = c(loss, val_loss), names_to = "series", values_to = "loss")
hist_long$series <- factor(hist_long$series, levels = c("loss", "val_loss"),
                           labels = c("Train loss", "Val loss"))

ggplot(hist_long, aes(x = epoch, y = loss, color = series)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "LSTM training and validation loss",
    x = "Epoch",
    y = "Loss (MSE)",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 14-day forecast

```{r}
last_seq <- values[(test_start - SEQ_LEN):(test_start - 1), , drop = FALSE]
last_seq_scaled <- scale_fun(last_seq)
forecast_scaled <- numeric(HORIZON)
current <- last_seq_scaled

for (h in seq_len(HORIZON)) {
  x_h <- array(current, dim = c(1, SEQ_LEN, 1))
  pred <- model$predict(x_h, verbose = 0L)
  forecast_scaled[h] <- as.numeric(pred)[1]
  current <- rbind(current[-1, , drop = FALSE], pred[1, 1])
}

forecast <- inv_scale_fun(forecast_scaled)
forecast_df <- data.frame(
  date = test_dates,
  actual = as.numeric(test_values),
  forecast = forecast
)
forecast_df
```

## Plot: actual vs LSTM forecast (14 days)

```{r}
library(scales)

ggplot(forecast_df, aes(x = date)) +
  geom_line(aes(y = actual, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = forecast, color = "LSTM forecast"), linewidth = 0.8) +
  geom_point(aes(y = actual, color = "Actual"), size = 2) +
  geom_point(aes(y = forecast, color = "LSTM forecast"), size = 2, shape = 17) +
  scale_color_manual(
    values = c("Actual" = "#332288", "LSTM forecast" = "#EE7733"),
    name = NULL
  ) +
  labs(
    title = "14-day forecast: actual vs LSTM",
    x = "Date",
    y = "Avg trip duration (minutes)"
  ) +
  scale_x_date(date_breaks = "2 days", date_labels = "%b %d") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Evaluation: sMAPE and MASE

```{r}
library(Metrics)
library(yardstick)

mae_train <- mean(abs(diff(as.numeric(train_values))))
# Metrics::smape returns proportion; * 100 for percentage
lstm_smape <- Metrics::smape(forecast_df$actual, forecast_df$forecast) * 100
lstm_mase <- yardstick::mase_vec(
  truth = forecast_df$actual,
  estimate = forecast_df$forecast,
  m = 1,
  mae_train = mae_train
)

cat("LSTM sMAPE (%):", round(lstm_smape, 4), "\n")
cat("LSTM MASE:", round(lstm_mase, 4), "\n")
```

## Comparison: ARIMA vs LSTM

```{r}
arima_path <- file.path(getwd(), "../data", "arima_forecast_14day.csv")
comparison <- data.frame(
  Model = "LSTM",
  sMAPE_pct = round(lstm_smape, 4),
  MASE = round(lstm_mase, 4)
)
if (file.exists(arima_path)) {
  arima_df <- read.csv(arima_path)
  arima_df$actual <- as.numeric(arima_df$actual)
  arima_df$forecast <- as.numeric(arima_df$forecast)
  arima_smape <- Metrics::smape(arima_df$actual, arima_df$forecast) * 100
  arima_mase <- yardstick::mase_vec(
    truth = arima_df$actual,
    estimate = arima_df$forecast,
    m = 1,
    mae_train = mae_train
  )
  comparison <- rbind(
    data.frame(Model = "ARIMA", sMAPE_pct = round(arima_smape, 4), MASE = round(arima_mase, 4)),
    comparison
  )
}
names(comparison)[2] <- "sMAPE (%)"
print(comparison)
```

## Time series plot (full year, same style as finalproject.qmd)

```{r}
ts_plot_df <- data.frame(
  Date = data$pickup_date,
  Duration = data$avg_duration_min
)

ggplot(ts_plot_df, aes(x = Date, y = Duration)) +
  geom_line() +
  labs(
    title = "Daily Average Uber Trip Duration (2024)",
    x = "Month",
    y = "Average Duration (minutes)"
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b"
  ) +
  theme_minimal()
```

## Save forecast for comparison with ARIMA

```{r}
out_path <- file.path(getwd(), "../data", "lstm_forecast_14day.csv")
write.csv(forecast_df, out_path, row.names = FALSE)
```
